Strength, mating success, and immune and nutritional costs in a population sample of US women and men: a registered report

### 2024-09-23 paper.qmd
Grammar fixes. Changed 'lifetime number of partners' to 'lifetime sexual partners' and 'past year number of partners' to 'past-year sexual partners' throughout.

Theory and evidence suggest that the mating benefits of muscle mass in human males trade off with costs of increased energy intake and decreased measures of innate immunity, likely due to an evolutionary history of sexual selection. Lassek & Gaulin (2009) demonstrated a positive association between male fat free mass and limb muscle volume and mating success, but did not investigate women. It is therefore unknown if females experience a similar tradeoff. Using data from the 2013-2014 phase of the National Health and Nutrition Examination Survey (NHANES), a large nationally representative sample of the US adults (N = `{r} nrow(designsH2$d.design.adults)`), we tested the prediction from the sexual selection hypothesis that the association of upper body strength, proxied by grip strength, with mating success is significantly positive for males and significantly less so for females. We found a main effect of strength on mating success proxied by lifetime number of sexual partners and current partnered status, but not past year number of sexual partners or age at first intercourse. We found consistent evidence for a $grip\ strength\times sex$ interaction on partnered status, such that strength was significantly more important for male partnered status than female (but no significant interaction for lifetime number of partners). We also tested for tradeoffs of upper body strength with immune and dietary intake and found a positive relationship between grip strength and protein and energy intake, but no significant relationship between grip strength and innate immune function. Our results suggest that sexually dimorphic upper body strength might have evolved, in part, by increasing male long-term mating success.

@@

Theoretically predicted relationships between upper body strength and various indices of mating success, such as numbers of partners, partnered status, and age at first sex, for both men and women. According to the sexual selection hypothesis, sexual dimorphism in upper body strength is explained by a positive effect on male mating success that is greater than its effect on female mating success, i.e., a sex:strength interaction (the hypothesis does allow some positive or negative relationship between female strength and mating success, however, which is represented by the grey uncertainty band on the female slope).
:::

@@

In order to refine our hypotheses and statistical models for our confirmatory study, which will use data we have not yet observed, we first conducted a pilot study whose results we report here.

@@

The pilot study uses the 2011-2012 dataset, whereas the 2013-2014 dataset has been held out for confirmatory analysis pending in principle acceptance of this registered report. Grip strength, our proxy for physical formidability and our key predictor variable, was only collected in these years [@smith2022 investigated the association of grip strength with depression in the 2013-14 data, but did not investigate or observe the relationship between grip strength and any of the mating success outcomes described below]. We include data from US adults between the ages of 18 and 60, the years when grip strength is the most stable [@hogrel2015]. For each model, participants will be included if they have complete data for each predictor and outcome variable.

@@

Our main outcome was mating success. In industrialized populations like the U.S., widespread access to contraceptives uncouples reproductive success from sexual behavior. Measures of mating success, including number of sexual partners and age at first sexual intercourse, are used as proxies as they are assumed to have been strongly correlated with male reproductive success under ancestral conditions [@perusse1993]. We operationalized mating success in three ways following @lassek2009: 1) Total sexual partners was a count based on responses to the question: "In your lifetime, with how many men/women have you had any kind of sex?" 2) Total sexual partners in the past year, from the question: "In the past 12 months, with how many men/women have you had any kind of sex?" Because of the way these questions are framed, (e.g. female participants were asked about their male partners and vice versa) these variables represent heterosexual partners (however, it is not necessarily the case that all of the participants included in these models identified as heterosexual). 3) Age at first sexual intercourse reported from: "How old were you when you had sex for the first time?" The sexual behavior questionnaire was self-administered on a computer in a private room at the examination center, using the Audio Computer Assisted Self Interview system, which allows participants to hear questions through headphones as well as read them on screen. Only respondents who could self-report were asked these questions.

Lifetime and past year numbers of sexual partners index success through short term mating strategies. Lifetime numbers are probably a more precise proxy of the use of a short-term mating strategy as these are integrated over the entire life, whereas past year numbers likely fluctuate depending on partnered status and other transient factors. However, both are somewhat ambiguous measures as they could also represent repeated rejection, or a tendency to exaggerate or understate the actual numbers. Neither clearly indexes female reproductive success because multiple sexual partners are likely to have increased male fitness more than female fitness [i.e., the Bateman gradient, @Lehtonen_2022]. We therefore also operationalized mating success in a fourth outcome measure as partnered status as a proxy for a long-term mating strategy that, as discussed earlier, benefits both sexes. Partnered status included participants who reported either being married or living with a partner whereas unpartnered included participants who reported being single, widowed, divorced, or separated [@lassek2009, included marital status as a control variable but not as an outcome variable].

@@

Our main explanatory variable of mating success is formidability, which @lassek2009 operationalized as fat-free mass and limb muscle volume. Since those variables were not available in our data, we operationalized formidability as combined grip strength (kg), the sum of the highest of three readings taken on each hand using a dynamometer (hereafter 'grip strength'). All regression models also included sex (male/female), our second main explanatory variable. All models also controlled for age in years and partnered status (not included in models where partnered status was the outcome variable).

The interaction between sex and grip strength was crucial for our models since we want to test if the effect of grip strength on mating and immune outcomes is different for women than for men. Grip strength is highly sexually dimorphic, however, with about 90% of men stronger than about 90% of women [@smith2022], and is thus highly collinear with sex, a problem for regression modeling and interpretation. We therefore computed a sex-specific grip strength value as follows: for each sex we centered grip strength at the sex-specific mean and divided by two times the sex-specific standard deviation. Thus, for women, low or high values are relative to other women, and for men, low or high values are relative to other men. We denote the use of this predictor in text as 'sex-specific grip strength' and figures below as 'Grip Strength (S by sex).' The sex-specific values were used in all models except the immune function and dietary intake models described below, which assess the costs of high upper body strength. Here, an index of absolute rather than relative strength is called for, and we used grip strength standardized across both sexes, denoted as Grip Strength (S) in results below. (See the post hoc analyses section, where in addition to grip strength, we also fit models using lean body masses.)

We also interacted sex with age and partnered status. We interacted age with sex since menopause occurs within the age range of our participants. For models of past year number of sexual partners we included an interaction between partnered status and sex-specific grip strength since being currently partnered would plausibly impact mating behavior in the past year regardless of strength. Continuous-valued predictor variables were centered and standardized by 2 standard deviations which approximately matches the variation in binary variables like sex, thus making the regression coefficients more comparable [@gelman2008].

@@

We attempted to replicate the regression models in @lassek2009 as closely as possible. There were some unavoidable differences, however. First, our study aimed to investigate the role of sex; thus we include sex as a main effect as well as in interactions with key predictor variables. Second, some variables included in @lassek2009 were not available in our data. Third, we included additional control variables as noted below. Finally, we treated our pilot study as an exploratory study in which we fit numerous models not reported here. The models we report in the pilot study and which we will test exactly in the confirmatory study represent those that we consider to best test the hypotheses in @lassek2009 based on the theories we described above, the results of our exploratory analyses, and our constraints.

@@

Our modeling strategy takes inspiration from @lassek2009. However, @lassek2009 utilized stepwise regression to automatically eliminate candidate predictors from their models which has since been found to overfit data and therefore estimated coefficients fail to replicate in future samples [@smith2018a]. Instead, we will fit generalized linear regression models (GLMs) with prespecified treatment and control variables chosen based on theoretical considerations and our exploratory analyses. Control variables included a much wider range of theoretically motivated potential confounds than used by @lassek2009, organized into themes. We specified five models for each mating success outcome measure in order to determine if the effect of sex-specific grip strength on those outcomes was due to confounds with socioeconomic, health, hormone, or physical activity variables that have been associated with strength and sexual behavior.

@@

We first derived a simple model based on that reported in Table 2 in @lassek2009 for each outcome measure (total number of partners, past year number of partners, and age at first intercourse) predicted by age, sex, sex-specific grip strength, and partnered status. This model also includes an anthropometric control, body mass index (BMI), calculated as $kg/m^2$, since body size could impact either strength or mating outcomes. We interacted age, sex-specific grip strength, and BMI with sex to assess differences in these predictors for men and women.

@@

All analyses were completed in `{r} R.version.string`, using the survey package [`{r} packageVersion("survey")`, @lumley2021] in order to incorporate the survey sampling weights and to preserve the representative structure of the sample. We modeled lifetime number of partners, past year numbers of partners, and white blood cell count using a quasi-Poisson generalized linear regression model (glm), since these variables are overdispersed count data. We modeled age at first sexual intercourse and energy and protein intakes with a Gaussian glm. We modeled partnered status using a binomial glm. Analyses included adults ages 18-60 because this is the period when the majority of reproduction takes place and because these are the years when grip strength is the most stable.

@@

# Last year
female_median_year <- 
  svyquantile(~sex_partners_year, designsG$d.design.adult.female, 0.5, na.rm = T)$sex_partners_year[1,1]
male_median_year <- 
  svyquantile(~sex_partners_year, designsG$d.design.adult.male, 0.5, na.rm = T)$sex_partners_year[1,1]
high_partners_year <- sum(d_G$sex_partners_year >= 10, na.rm=T)
```

Our initial analyses found little support for a sex difference in the effect of sex-specific grip strength on mating success, contrary to the sexual selection hypothesis. We traced this failure to the unusually high number of lifetime sex partners reported by some men and women. We discovered that whereas women reported a median number of `{r} female_median` lifetime sex partners, and men `{r} male_median` partners, there were `{r} high_partners` individuals who reported 100 or more partners, a few of whom reported 1000 or more (@fig-ecdf). We decided to restrict our sample to individuals with fewer than 100 lifetime sex partners for several reasons. First, support for the sexual selection hypothesis was weak if these individuals were included, but much stronger if they were excluded. We suspect that because our sample is large and nationally representative, it probably includes sex workers, as well as individuals who regularly use the services of sex workers, and whose high partner numbers would therefore not reflect mate attraction or intrasexual competition. Second, there was an unusually high number of participants reporting exactly 100 lifetime sex partners, suggesting inaccurate recollections. Finally, our statistical models do not easily accommodate the high degree of overdispersion caused by these few exceptionally high partner numbers, and our extensive simulations showed that high overdispersion would reduce our statistical power to detect sex differences. We applied this sample cutoff for all models of mating success. For our tradeoff models of innate immunity, energy intake, and protein intake we did not restrict our sample based on reported number of lifetime partners. We used the same cutoff for our confirmatory study.

```{r}
#| label: fig-ecdf
#| fig-width: 10
#| fig-cap: Cumulative distribution of the number of sex partners, by sex. Participants who reported 100 or more partners were removed from the analysis. The x-axis is on a log scale, so 1 was added to the number of partners to prevent removal of individuals with 0 sex partners (only for this Figure and not for any analyses).

@@

Our pilot results revealed strong positive significant main effects of sex-specific grip strength on partnered status, numbers of lifetime partners, and age at first sex after controlling for all five sets of potential confounds. The only exception was that the sex-specific grip strength coefficient for age at first sex was not significant after controlling for socioeconomic factors, although the sign and magnitude was very similar to those of these coefficients in the models with other control variables. There was no significant effect of sex-specific grip strength on the number of partners in the past year. See the left column of @fig-coefs.

There were also clear sex differences in the effects of sex-specific grip strength on two mating success outcomes---partnered status and lifetime number of partners---with smaller effects for women compared to men, as predicted by the sexual selection hypothesis. There was also a persistent negative association of numbers of lifetime partners on the probability of partnered status in all models of this outcome (@sfig-tradeoff). Finally there was no evidence for a sex difference in the effects of sex-specific grip strength on past year number of partners or age at first sex, contrary to the sexual selection hypothesis. See the right column of @fig-coefs.

@@

For ease of comparison, pilot results discussed here are displayed in conjunction with confirmatory results in the Confirmatory Results section below.

Using sex-specific grip strength as a proxy for muscularity, this pilot study replicated findings from @lassek2009 that muscularity is significantly positively associated with two of the four indices of mating success, including our additional index of mating success---partnered status---even after controlling for a wider range of potential confounds including socioeconomic, health, hormone, and physical activity variables. The exceptions were that sex-specific grip strength was not a significant predictor of past year partners, nor of age at first sex in the socioeconomic control model; the coefficient of the latter, however, was similar in sign and magnitude to the significant sex-specific grip strength coefficients in the other models. Grip strength was also associated with immune, energy, and protein costs in the replication models, but not in the expanded control models of immune or energy costs; it remained a significant predictor of protein intake, however. The expanded control models included height, which might be a proxy for aspects of strength, such as lower body strength, that are not accounted for by grip strength.

The sexual selection hypothesis predicts that the effect of strength on mating success will be greater for men than for women, and this prediction was exceptionally well-supported for partnered status and numbers of lifetime partners. In humans, biological fitness depends critically on extensive investment in offspring which typically takes place in long-term partnerships. We therefore propose that the sexual selection hypothesis for the sexual dimorphism in physical formidability applies to long-term partnerships (proxied by our partnered variable), as well as short-term matings (proxied by partner numbers). Thus, our pilot study supports the hypothesis that muscularity increases both short- and long-term mating success for men more than it does for women. Number of lifetime partners was also a negative predictor of partnered status (@sfig-tradeoff), suggesting a tradeoff in pursuing these two reproductive strategies.

@@

Sex-specific grip strength was also not significantly associated with the number of partners in the last year, despite controlling for partnered status, which was a significant negative predictor. Approximately 60% of the sample was partnered (@stbl-stats), however, and stronger individuals, especially men, were more likely to be partnered. The pool of unpartnered individuals who might pursue multiple mating opportunities was therefore smaller and skewed toward lower grip strength individuals, and variation in this outcome over the relatively short time period (1 year) was low. Nevertheless, the non-significant sex-specific grip strength coefficients were all positive, raising the possibility that a higher powered study might successfully detect a positive effect of sex-specific grip strength on number of partners in the last year.

@@

In a large, nationally representative US sample, we predict that formidability will be significantly positively associated with mating success as indexed by partnered status and lifetime number of sexual partners, but not last year number of partners, nor age at first sex after controlling for ethnicity. We also predict a significant interaction with sex such that sex-specific grip strength will be a stronger predictor of male than female partnered status and number of lifetime partners, but not number of last year partners nor age at first sex. We also predict that there will be a significant negative association between formidability and immune function. Finally, we predict that formidability will be positively associated with dietary protein intake for both males and females.

@@

tribble(
              ~Outcome, ~`Grip Strength Coefficient`, ~`Grip Strength X Sex Coefficient`,
    "Partnered status",     "+",             "-",       
   "Lifetime partners",     "+",             "-",       
  "Past year partners",     "ns",            "ns",       
    "Age at first sex",     "ns",            "ns",       
            "Immunity",     "-",             "ns",       
              "Energy",     "ns",             "",       
             "Protein",     "+",              "",       
  ) |>    
  tt() |> 
  style_tt(j=c(1), align = 'l') 
```
## Power

Effects in replication studies are often smaller than in the original studies [@klein2018]. We therefore used simulations to estimate our power to detect grip strength and $grip\ strength\times sex$ coefficients that are 25-100% as large as the ones observed in our pilot study for partnered and lifetime partners, with a sample size equal to that in our unobserved data. We have high power to detect effects at least 75% as large as the ones observed in our pilot study, and reasonable power to detect those that are at least 50% as large (@stbl-power).

@@

Our confirmatory results revealed a strong positive significant main effect of sex-specific grip strength on partnered status and numbers of lifetime partners after controlling for all five sets of potential confounds. We found mixed evidence for a main effect of sex-specific grip strength in the age of first sex models, as it was not a significant predictor after controlling for socioeconomic, anthropometric, or physical activity controls. Strength was not a predictor of past year number of sexual partners in any control model.

Unlike findings from the pilot study, a clear sex difference in the effects of sex-specific grip strength only emerged for one mating success outcome---partnered status---with smaller effects for women compared to men, as predicted by the sexual selection hypothesis. The interaction between sex and sex-specific grip strength on lifetime number of sexual partners was not replicated in the confirmatory data.  Finally there was no evidence for a sex difference in the effects of sex-specific grip strength on past year number of partners or age at first sex, contrary to the sexual selection hypothesis. See @fig-coefs for coefficients of sex-specific grip strength and $grip\ strength\times sex$ interaction for each proxy of mating success in both pilot and confirmatory data. See @fig-allcoefs for coefficients of potential confounds in each model in the confirmatory data. For effects plots of the effect of sex-specific grip strength on each mating outcome in pilot and confirmatory studies (on the response scale), see @fig-effects.

@@

tribble(
              ~Outcome, ~Prediction, ~Supported, ~Prediction, ~Supported,
    "Partnered status",     "+",       "✓",       "-",       "✓",
   "Lifetime partners",     "+",       "✓",       "-",       "❌",
  "Past year partners",     "ns",      "✓",       "ns",       "✓",
    "Age at first sex",     "ns",      "✓",       "ns",       "✓",
            "Immunity",     "-",       "❌*",     "ns",       "✓",
              "Energy",     "ns",      "❌*",      "",        "",
             "Protein",     "+",       "✓",        "",        ""
  ) |>      
  tt() |> 
  style_tt(j=c(1,3,5), align = 'l') |> 
  group_tt(j = list("Grip strength coefficient" = 2:3, "Grip strength X sex coefficient" = 4:5))
```

@@

Our post hoc analyses revealed that coefficients for sex-specific arm lean mass were generally similar in direction to those of sex-specific grip strength discussed above, including a strong positive significant main effect on partnered status and numbers of lifetime partners after controlling for all five sets of potential confounds. In addition, sex-specific arm lean mass had a significantly smaller association with partnered status for women than men, consistent with the sexual selection hypothesis. We found mixed evidence for a main effect of sex-specific lean mass in the age of first sex models, as these were not significant predictors after controlling for socioeconomic or physical activity controls. As with sex-specific grip strength, sex-specific arm lean mass was not a predictor of past year number of sexual partners in any control model. Sex-specific leg, trunk, and total lean mass were not consistently related to any mating success outcome. See @fig-leancoefs and @sfig-legs - @sfig-total.

@@

In the Stage 2 Confirmatory Study, we found consistent support for a positive relationship between both sex-specific combined grip strength and sex-specific arm lean mass, and two measures of mating success: being currently partnered, and lifetime number of sexual partners for men, controlling for numerous possible confounds. We found no consistent association between either sex-specific combined grip strength or any sex-specific lean masses and age at first sex or number of past year sexual partners.

@@

### The negative association of numbers of lifetime sexual partners with partnered status

>>>>

Theory and evidence suggest that the mating benefits of muscle mass in human males trade off with costs of increased energy intake and decreased measures of innate immunity, likely due to an evolutionary history of sexual selection. Lassek & Gaulin (2009) demonstrated a positive association between male fat free mass and limb muscle volume and mating success, but did not investigate women. It is therefore unknown if females experience a similar tradeoff. Using data from the 2013-2014 phase of the National Health and Nutrition Examination Survey (NHANES), a large nationally representative sample of US adults (N = `{r} nrow(designsH2$d.design.adults)`), we tested the prediction from the sexual selection hypothesis that the association of upper body strength, proxied by grip strength, with mating success is significantly positive for males and significantly less so for females. We found a main effect of strength on mating success proxied by lifetime number of sexual partners and current partnered status, but not past-year number of sexual partners or age at first intercourse. We found consistent evidence for a $grip\ strength\times sex$ interaction on partnered status, such that strength was significantly more important for male partnered status than female (but no significant interaction for lifetime sexual partners). We also tested for tradeoffs of upper body strength with immune and dietary intake and found a positive relationship between grip strength and protein and energy intake, but no significant relationship between grip strength and innate immune function. Our results suggest that sexually dimorphic upper body strength might have evolved, in part, by increasing male long-term mating success.

@@

Theoretically predicted relationships between upper body strength and various indices of mating success, such as numbers of sexual partners, partnered status, and age at first sex, for both men and women. According to the sexual selection hypothesis, sexual dimorphism in upper body strength is explained by a positive effect on male mating success that is greater than its effect on female mating success, i.e., a sex x strength interaction (the hypothesis does allow for some degree of positive or negative relationship between female strength and mating success, however, which is represented by the grey uncertainty band on the female slope).
:::

@@

In order to refine our hypotheses and statistical models for our confirmatory study, we first conducted a pilot study whose results we report here.

@@

The pilot study used the 2011-2012 dataset, whereas the 2013-2014 dataset was held out for confirmatory analysis pending in principle acceptance of this registered report. Grip strength, our proxy for physical formidability and our key predictor variable, was only collected in these years [@smith2022 investigated the association of grip strength with depression in the 2013-14 data, but did not investigate or observe the relationship between grip strength and any of the mating success outcomes described below]. We include data from US adults between the ages of 18 and 60, the years when grip strength is the most stable [@hogrel2015]. For each model, participants were included if they had complete data for each predictor and outcome variable.

@@

Our main outcome was mating success. In industrialized populations like the U.S., widespread access to contraceptives uncouples reproductive success from sexual behavior. Measures of mating success, including number of sexual partners and age at first sexual intercourse, are used as proxies as they are assumed to have been strongly correlated with male reproductive success under ancestral conditions [@perusse1993]. We operationalized mating success in three ways following @lassek2009: 1) Lifetime sexual partners was a count based on responses to the question: "In your lifetime, with how many men/women have you had any kind of sex?" 2) Past-year sexual partners, from the question: "In the past 12 months, with how many men/women have you had any kind of sex?" Because of the way these questions are framed, (e.g. female participants were asked about their male partners and vice versa) these variables represent heterosexual partners (however, it is not necessarily the case that all of the participants included in these models identified as heterosexual). 3) Age at first sexual intercourse reported from: "How old were you when you had sex for the first time?" The sexual behavior questionnaire was self-administered on a computer in a private room at the examination center, using the Audio Computer Assisted Self Interview system, which allows participants to hear questions through headphones as well as read them on screen. Only respondents who could self-report were asked these questions.

Lifetime and past-year numbers of sexual partners index success through short term mating strategies. Lifetime numbers are probably a more precise proxy of the use of a short-term mating strategy as these are integrated over the entire life, whereas past-year numbers likely fluctuate depending on partnered status and other transient factors. However, both are somewhat ambiguous measures as they could also represent repeated rejection, or a tendency to exaggerate or understate the actual numbers. Neither clearly indexes female reproductive success because multiple sexual partners are likely to have increased male fitness more than female fitness [i.e., the Bateman gradient, @Lehtonen_2022]. We therefore also operationalized mating success in a fourth outcome measure as partnered status as a proxy for a long-term mating strategy that, as discussed earlier, benefits both sexes. Partnered status included participants who reported either being married or living with a partner whereas unpartnered included participants who reported being single, widowed, divorced, or separated [@lassek2009, included marital status as a control variable but not as an outcome variable].

@@

Our main explanatory variable of mating success is formidability, which @lassek2009 operationalized as fat-free mass and limb muscle volume. Since those variables were not available in our data, we operationalized formidability as combined grip strength (kg), the sum of the highest of three readings taken on each hand using a dynamometer (hereafter 'grip strength'). All regression models also included sex (male/female), our second main explanatory variable. All models also controlled for age in years and partnered status (not included in models where partnered status was the outcome variable). Continuous-valued predictor variables were centered and standardized by 2 standard deviations which approximately matches the variation in binary variables like sex, thus making the regression coefficients more comparable [@gelman2008]. Standardized variables are denoted in figures as 'Variable Name (S)'.

The interaction between sex and grip strength was crucial for our models since we wanted to test if the effect of grip strength on mating and immune outcomes is different for women than for men. Grip strength is highly sexually dimorphic, however, with about 90% of men stronger than about 90% of women [@smith2022], and is thus highly collinear with sex, a problem for regression modeling and interpretation. We therefore computed a sex-specific grip strength value as follows: for each sex we centered grip strength at the sex-specific mean and divided by two times the sex-specific standard deviation. Thus, for women, low or high values are relative to other women, and for men, low or high values are relative to other men. We denote the use of this predictor in text as 'sex-specific grip strength' and figures below as 'Grip Strength (S by sex).' The sex-specific values were used in all models except the immune function and dietary intake models described below, which assess the costs of high upper body strength. Here, an index of absolute rather than relative strength is called for, and we used grip strength standardized across both sexes, denoted as Grip Strength (S) in results below. (See the post hoc analyses section, where in addition to grip strength, we also fit models using lean body masses.)

We also interacted sex with age and partnered status. We interacted age with sex since menopause occurs within the age range of our participants. For models of past-year sexual partners we included an interaction between partnered status and sex-specific grip strength since being currently partnered would plausibly impact mating behavior in the past year regardless of strength.

@@

We attempted to replicate the regression models in @lassek2009 as closely as possible. There were some unavoidable differences, however. First, our study aimed to investigate the role of sex; thus we include sex as a main effect as well as in interactions with key predictor variables. Second, some variables included in @lassek2009 were not available in our data. Third, we included additional control variables as noted below. Finally, we treated our pilot study as an exploratory study in which we fit numerous models not reported here. The models we report in the pilot study and which we then tested exactly in the confirmatory study represent those that we consider to best test the hypotheses in @lassek2009 based on the theories we described above, the results of our exploratory analyses, and our constraints.

@@

Our modeling strategy takes inspiration from @lassek2009. However, @lassek2009 utilized stepwise regression to automatically eliminate candidate predictors from their models which has since been found to overfit data and therefore estimated coefficients fail to replicate in future samples [@smith2018a]. Instead, we fit generalized linear regression models (GLMs) with prespecified treatment and control variables chosen based on theoretical considerations and our exploratory analyses. Control variables included a much wider range of theoretically motivated potential confounds than used by @lassek2009, organized into themes. We specified five models for each mating success outcome measure in order to determine if the effect of sex-specific grip strength on those outcomes was due to confounds with socioeconomic, health, hormone, or physical activity variables that have been associated with strength and sexual behavior.

@@

We first derived a simple model based on that reported in Table 2 in @lassek2009 for each outcome measure (lifetime sexual partners, past-year sexual partners, and age at first intercourse) predicted by age, sex, sex-specific grip strength, and partnered status. This model also includes an anthropometric control, body mass index (BMI), calculated as $kg/m^2$, since body size could impact either strength or mating outcomes. We interacted age, sex-specific grip strength, and BMI with sex to assess differences in these predictors for men and women.

@@

All analyses were completed in `{r} R.version.string`, using the survey package [`{r} packageVersion("survey")`, @lumley2021] in order to incorporate the survey sampling weights and to preserve the representative structure of the sample. We modeled lifetime sexual partners, past-year sexual partners, and white blood cell count using a quasi-Poisson generalized linear regression model (glm), since these variables are overdispersed count data. We modeled age at first sexual intercourse and energy and protein intakes with a Gaussian glm. We modeled partnered status using a binomial glm. Analyses included adults ages 18-60 because this is the period when the majority of reproduction takes place and because these are the years when grip strength is the most stable.

@@

# Last year
female_median_year <- 
  svyquantile(~sex_partners_year, designsG$d.design.adult.female, 0.5, na.rm = T)$sex_partners_year[1,1]
male_median_year <- 
  svyquantile(~sex_partners_year, designsG$d.design.adult.male, 0.5, na.rm = T)$sex_partners_year[1,1]
high_partners_year <- sum(d_G$sex_partners_year >= 10, na.rm=T)
```
For ease of comparison, pilot results described here are displayed in conjunction with confirmatory results in the Confirmatory Results section below.

Our initial analyses found little support for a sex difference in the effect of sex-specific grip strength on mating success, contrary to the sexual selection hypothesis. We traced this failure to the unusually high number of lifetime sexual partners reported by some men and women. We discovered that whereas women reported a median number of `{r} female_median` lifetime sexual partners, and men `{r} male_median` partners, there were `{r} high_partners` individuals who reported 100 or more partners, a few of whom reported 1000 or more (@fig-ecdf). We decided to restrict our sample to individuals with fewer than 100 lifetime sexual partners for several reasons. First, support for the sexual selection hypothesis was weak if these individuals were included, but much stronger if they were excluded. We suspect that because our sample is large and nationally representative, it probably includes sex workers, as well as individuals who regularly use the services of sex workers, and whose high partner numbers would therefore not reflect mate attraction or intrasexual competition. Second, there was an unusually high number of participants reporting exactly 100 lifetime sexual partners, suggesting inaccurate recollections. Finally, our statistical models do not easily accommodate the high degree of overdispersion caused by these few exceptionally high partner numbers, and our extensive simulations showed that high overdispersion would reduce our statistical power to detect sex differences. We applied this sample cutoff for all models of mating success. For our tradeoff models of innate immunity, energy intake, and protein intake we did not restrict our sample based on reported number of lifetime partners. We used the same cutoff for our confirmatory study.

```{r}
#| label: fig-ecdf
#| fig-width: 10
#| fig-cap: Cumulative distribution of the number of sexual partners, by sex. Participants who reported 100 or more partners were removed from the analysis. The x-axis is on a log scale, so 1 was added to the number of partners to prevent removal of individuals with 0 sexual partners (only for this Figure and not for any analyses).

@@

Our pilot results revealed strong positive significant main effects of sex-specific grip strength on partnered status, lifetime sexual partners, and age at first sex after controlling for all five sets of potential confounds. The only exception was that the sex-specific grip strength coefficient for age at first sex was not significant after controlling for socioeconomic factors, although the sign and magnitude was very similar to those of these coefficients in the models with other control variables. There was no significant effect of sex-specific grip strength on the number of sexual partners in the past year. See the left column of @fig-coefs.

There were also clear sex differences in the effects of sex-specific grip strength on two mating success outcomes---partnered status and lifetime sexual partners---with smaller effects for women compared to men, as predicted by the sexual selection hypothesis. There was also a persistent negative association of lifetime sexual partners on the probability of partnered status in all models of this outcome (@sfig-tradeoff). Finally there was no evidence for a sex difference in the effects of sex-specific grip strength on past-year sexual partners or age at first sex, contrary to the sexual selection hypothesis. See the right column of @fig-coefs.

@@

Using sex-specific grip strength as a proxy for muscularity, this pilot study replicated findings from @lassek2009 that muscularity is significantly positively associated with two of the four indices of mating success, including our additional index of mating success---partnered status---even after controlling for a wider range of potential confounds including socioeconomic, health, hormone, and physical activity variables. The exceptions were that sex-specific grip strength was not a significant predictor of past-year partners, nor of age at first sex in the socioeconomic control model; the coefficient of the latter, however, was similar in sign and magnitude to the significant sex-specific grip strength coefficients in the other models. Grip strength was also associated with immune, energy, and protein costs in the Lassek and Gaulin models, but not in the expanded control models of immune or energy costs; it remained a significant predictor of protein intake, however. The expanded control models included height, which might be a proxy for aspects of strength, such as lower body strength, that are not accounted for by grip strength.

The sexual selection hypothesis predicts that the effect of strength on mating success will be greater for men than for women, and this prediction was exceptionally well-supported for partnered status and lifetime sexual partners. In humans, biological fitness depends critically on extensive investment in offspring which typically takes place in long-term partnerships. We therefore propose that the sexual selection hypothesis for the sexual dimorphism in physical formidability applies to long-term partnerships (proxied by our partnered variable), as well as short-term matings (proxied by partner numbers). Thus, our pilot study supports the hypothesis that muscularity increases both short- and long-term mating success for men more than it does for women. Lifetime sexual partners was also a negative predictor of partnered status (@sfig-tradeoff), suggesting a tradeoff in pursuing these two reproductive strategies.

@@

Sex-specific grip strength was also not significantly associated with the number of sexual partners in the past year, despite controlling for partnered status, which was a significant negative predictor. Approximately 60% of the sample was partnered (@stbl-stats), however, and stronger individuals, especially men, were more likely to be partnered. The pool of unpartnered individuals who might pursue multiple mating opportunities was therefore smaller and skewed toward lower grip strength individuals, and variation in this outcome over the relatively short time period (1 year) was low. Nevertheless, the non-significant sex-specific grip strength coefficients were all positive, raising the possibility that a higher powered study might successfully detect a positive effect of sex-specific grip strength on number of partners in the past year.

@@

In a large, nationally representative US sample, we predict that formidability will be significantly positively associated with mating success as indexed by partnered status and lifetime number of sexual partners, but not past-year number of partners, nor age at first sex after controlling for ethnicity. We also predict a significant interaction with sex such that sex-specific grip strength will be a stronger predictor of male than female partnered status and lifetime number of sexual partners, but not number of past-year sexual partners nor age at first sex. We also predict that there will be a significant negative association between formidability and immune function. Finally, we predict that formidability will be positively associated with dietary protein intake for both males and females.

@@

tribble(
              ~Outcome, ~`Grip Strength Coefficient`, ~`Grip Strength X Sex Coefficient`,
    "Partnered status",     "+",             "-",       
   "Lifetime partners",     "+",             "-",       
  "Past-year partners",     "ns",            "ns",       
    "Age at first sex",     "ns",            "ns",       
            "Immunity",     "-",             "ns",       
              "Energy",     "ns",             "",       
             "Protein",     "+",              "",       
  ) |>    
  tt() |> 
  style_tt(j=c(1), align = 'l') 
```
## Power

Effects in replication studies are often smaller than in the original studies [@klein2018]. We therefore used simulations to estimate our power to detect grip strength and $grip\ strength\times sex$ coefficients that are 25-100% as large as the ones observed in our pilot study for partnered and lifetime sexual partners, with a sample size equal to that in our unobserved data. We have high power to detect effects at least 75% as large as the ones observed in our pilot study, and reasonable power to detect those that are at least 50% as large (@stbl-power).

@@

Our confirmatory results revealed a strong positive significant main effect of sex-specific grip strength on partnered status and lifetime number of partners after controlling for all five sets of potential confounds. We found mixed evidence for a main effect of sex-specific grip strength in the age of first sex models, as it was not a significant predictor after controlling for socioeconomic, anthropometric, or physical activity controls. Grip strength was not a predictor of past-year number of sexual partners in any control model.

Unlike findings from the pilot study, a clear sex difference in the effects of sex-specific grip strength only emerged for one mating success outcome---partnered status---with smaller effects for women compared to men, as predicted by the sexual selection hypothesis. The interaction between sex and sex-specific grip strength on lifetime number of sexual partners was not replicated in the confirmatory data.  Finally there was no evidence for a sex difference in the effects of sex-specific grip strength on past-year number of partners or age at first sex, contrary to the sexual selection hypothesis. See @fig-coefs for coefficients of sex-specific grip strength and $grip\ strength\times sex$ interaction for each proxy of mating success in both pilot and confirmatory data. See @fig-allcoefs for coefficients of potential confounds in each model in the confirmatory data. For effects plots of the effect of sex-specific grip strength on each mating outcome in pilot and confirmatory studies (on the response scale), see @fig-effects.

@@

tribble(
              ~Outcome, ~Prediction, ~Supported, ~Prediction, ~Supported,
    "Partnered status",     "+",       "✓",       "-",       "✓",
   "Lifetime partners",     "+",       "✓",       "-",       "❌",
  "Past-year partners",     "ns",      "✓",       "ns",       "✓",
    "Age at first sex",     "ns",      "✓",       "ns",       "✓",
            "Immunity",     "-",       "❌*",     "ns",       "✓",
              "Energy",     "ns",      "❌*",      "",        "",
             "Protein",     "+",       "✓",        "",        ""
  ) |>      
  tt() |> 
  style_tt(j=c(1,3,5), align = 'l') |> 
  group_tt(j = list("Grip strength coefficient" = 2:3, "Grip strength X sex coefficient" = 4:5))
```

@@

Our post hoc analyses revealed that coefficients for sex-specific arm lean mass were generally similar in direction to those of sex-specific grip strength discussed above, including a strong positive significant main effect on partnered status and lifetime number of partners after controlling for all five sets of potential confounds. In addition, sex-specific arm lean mass had a significantly smaller association with partnered status for women than men, consistent with the sexual selection hypothesis. We found mixed evidence for a main effect of sex-specific lean mass in the age of first sex models, as these were not significant predictors after controlling for socioeconomic or physical activity controls. As with sex-specific grip strength, sex-specific arm lean mass was not a predictor of past-year number of sexual partners in any control model. Sex-specific leg, trunk, and total lean mass were not consistently related to any mating success outcome. See @fig-leancoefs and @sfig-legs - @sfig-total.

@@

In the Stage 2 Confirmatory Study, we found consistent support for a positive relationship between both sex-specific combined grip strength and sex-specific arm lean mass, and two measures of mating success: being currently partnered, and lifetime number of sexual partners for men, controlling for numerous possible confounds. We found no consistent association between either sex-specific combined grip strength or any sex-specific lean masses and age at first sex or number of past-year sexual partners.

@@

### The negative association of lifetime number of sexual partners with partnered status
